Approximation as Differentiation

### 2025-07-10 main/models.tex
[ tex ] Some discussion of using Fam(LatGal)

The fact that morphisms in $\Fam(C)$ compose according to a chain rule
means that the categories we have seen so far embed into $\Fam(C)$ for
appropriate $C$. If we let $\FinVect$ be the category of finite
dimensional real vector spaces, then:

@@

\begin{definition}
  \item
  \begin{enumerate}
  \item The category $\LatGal$ has bounded lattices as objects and
    Galois connections as morphisms, with the right adjunct going in
    the ``forward'' direction.
  \item The category $\MeetSLat$ has meet-semilattices with top as
    objects and finite meet preserving functions as
    morphisms.
  \item The category $\JoinSLat$ has join-semilattices with bottom as
    objects and finite join preserving functions as
    morphisms.
  \end{enumerate}
\end{definition}

We can get an analogous result to \propref{embed-manifolds} for L-posets:

@@

The categories $\Fam(\cat{C})$ are the free coproduct completions of
categories $\cat{C}$, so they are guaranteed to have all coproducts:

@@

Using the infinitary coproducts and finite products, we can construct
a wide range of other useful semantic models of datatypes in
$\Fam(\cat{C})$. For example lists can be constructed as the coproduct
\begin{displaymath}
  \List(X) = \coprod_{n \in \mathbb{N}} X^n
\end{displaymath}
where $X^0 = 1$ (the terminal object) and $X^{n+1} = X \times X^n$.

Our category of interest, $\Fam(\LatGal)$ has coproducts and finite
products, because $\LatGal$ has products. As we shall see below, these
are also coproducts, which is essential to obtaining cartesian
closure.

@@

For cartesian closure of categories of families, we rely on the
following theorem of \citet{nunes2023}, specialised from their setting
with the general Grothendieck construction to $\Fam(\cat{C})$. This
relies on the definition of \emph{biproducts}, which we discuss below
in \secref{biproducts}.

\begin{theorem}[\cite{nunes2023}]
  \label{thm:fam-closed}
  \AGDA.  If $\cat{C}$ has biproducts and all products, then
  $\Fam(\cat{C})$ is cartesian closed\footnote{More precisely, if
    $\cat{C}$ has coproducts then we have a monoidal product on
    $\Fam(\cat{C})$ which is closed by this construction. When these
    coproducts are in fact biproducts, we get cartesian closure.}. On
  objects, the internal hom can be given by:
  \begin{displaymath}
    (X, \partial X) \to (Y, \partial Y) = (\Pi_{x : X}. \Sigma_{y : Y}. \cat{C}(\partial X(x), \partial Y(y)), \lambda f. \Pi_{x : X}. \partial Y(\pi_1(f\, x)))
  \end{displaymath}
\end{theorem}

Before we explore the hypotheses of this theorem, we give some
intuition on how it relates to differentiation. The Set-component of
$(X, \partial X) \to (Y, \partial Y)$ consists of exactly the
morphisms of $\Fam(\cat{C})$, rephrased into a single object. When
$\cat{C} = \FinVect$, these are functions with an associated linear
map at every point, and when $\cat{C} = \LatGal$, these are functions
with an associated Galois connection at every point. A tangent to a
function is then defined to be a mapping from points in the domain to
tangents in the codomain along the function.

Unfortunately, neither of the categories $\LatGal$ or $\FinVect$
satisfy the hypotheses of this theorem, because neither of them have
infinite products. We will consider ways to rectify this below in
\secref{fixing-completeness}.

\begin{remark}
  \label{rem:hermida-exponentials}
  There is another construction of internal homs on $\Fam(\cat{C})$
  arising from the use of fibrations for categorical logical relations
  proofs, due to \citet[Corollary 4.12]{hermida99}. If we assume that
  $\cat{C}$ is itself cartesian closed and has all products, then we
  could construct an internal hom as:
  \begin{displaymath}
    (X, \partial X) \to (Y, \partial Y) = (X \to Y, \lambda f. \Pi_{x : X}.\,\partial X(x) \to \partial Y(f\,x))
  \end{displaymath}
  However, for the purposes of modelling differentiable programs, this
  is fatally flawed in that neither of $\LatGal$ nor $\FinVect$ are
  cartesian closed, and there is no way of making them so without
  losing the property of having biproducts. We will use Hermida's
  construction in our definability proof in \secref{definability},
  where we use a logical relations argument to show that every
  morphism definable in the higher order language is also first-order
  definable.
\end{remark}

@@

Loosely stated, biproducts are objects that are both products and
coproducts. They can be defined in any category, as shown by
\citet{karvonen20}, but for our purposes it will be more convenient to
use the shorter definition in categories enriched in commutative
monoids:
\begin{definition}
  A category $\cat{C}$ is enriched in $\CMon$, the category of
  commutative monoids, if every homset $\cat{C}(X,Y)$ is a commutative
  monoid with $(+,0)$ and composition is bilinear:
  \begin{salign*}
    f \comp \zero = f = \zero \comp f
  \end{salign*}
  \begin{salign*}
    (f + g) \comp h &= (f \comp h) + (g \comp h) \\
    h \comp (f + g) &= (h \comp f) + (h \comp g)
  \end{salign*}
\end{definition}
In any $\CMon$-category we can define what it means to be the
biproduct of two objects:
\begin{definition}
  In a $\CMon$-category a biproduct is an object $X \biprod Y$
  together with morphisms

@@

\begin{example}
  The following categories are $\CMon$-enriched and have finite
  products, and hence biproducts:
  \begin{enumerate}
  \item In $\FinVect$, morphisms are linear maps and so can be added
    and have a zero map. Finite products are given by cartesian
    products of the underlying sets, with the vector operations
    defined pointwise.
  \item In $\LatGal$, right adjoints are summed using meets and left
    adjoints are summed using joins. The zero maps are given by the
    constantly $\top$ and constantly $\bot$ functions
    respectively. Products are given by the cartesian product of the
    underlying set.
  \item The $\LatGal$ example splits into two separate categories
    $\MeetSLat$ and $\JoinSLat$ of meet semilattices and join
    semilattices respectively, with their respective homomorphisms as
    the morphisms. As with $\LatGal$, these categories are
    $\CMon$-enriched and have finite products, and so finite
    biproducts.
  \end{enumerate}
\end{example}

The following proposition shows why the alternative construction of
exponentials in $\Fam(C)$ described in \remref{hermida-exponentials}
is not suitable:
\begin{proposition}
  A $\CMon$-category that is cartesian closed has trivial homsets:
  $\cat{C}(X,Y) \cong 1$.
\end{proposition}

\begin{proof}
  $\cat{C}(X, Y) \cong \cat{C}(0 + X,Y) \cong \cat{C}(0 \times X,Y)
  \cong \cat{C}(0,X \to Y) \cong 1$.
\end{proof}

@@

\begin{proposition}
  \label{prop:ho-embedding}
  The functor
  $F : \Fam(\LatGal) \to \Fam(\MeetSLat \times \JoinSLat^\op)$ is defined
  on objects as
  $F(X, \partial X) = (X, \lambda x. (\partial X(x), \partial
  X(x)))$. This functor is faithful and preserves coproducts and
  finite products.
\end{proposition}

With this embedding functor, we can see that the interpretation of
first-order types will be the same up to isomorphism in
$\Fam(\LatGal)$ and $\Fam(\MeetSLat \times \JoinSLat^\op)$, as long as
we interpret the base types as objects in $\Fam(\LatGal)$. At
higher-order, however, the meet-semilattice and join-semilattice
sides of the interpretation will diverge, and it is no longer clear
that the interpretation of programs using higher-order functions
internally will result in Galois connections. In \secref{definability}
we will see that every program with first-order type (even if it uses
higher-order functions internally) does in fact have an interpretation
definable in $\Fam(\LatGal)$.

% $\GLR(\Fam(\MeetSLat \times \JoinSLat^\op), F)$ fibred over
% $\Fam(\MeetSLat \times \JoinSLat^\op)$ with a full and faithful functor $\Fam(\LatGal) \to \GLR(\Fam(\MeetSLat \times \JoinSLat^\op), F)$.

@@

The above discussion has been somewhat abstract, so we now consider
some examples in the category $\Fam(\LatGal)$ and how they relate to
\GPS. We highlight how

@@

\bob{Move the discussion in \texttt{first-order.tex} to here, but
  construct the lifting monad in a more elementary way}

@@

\bob{flesh this out}

>>>>

The fact that morphisms in $\Fam(C)$ compose according to a chain rule
means that the categories we considered in \secref{approx-as-tangents}
embed into $\Fam(C)$ for appropriate $C$. If we let $\FinVect$ be the
category of finite dimensional real vector spaces and linear maps,
then:

@@

\begin{enumerate}
\item The category $\LatGal$ has bounded lattices as objects and
  Galois connections as morphisms, with the right adjunct going in
  the ``forward'' direction. The category $\Fam(\LatGal)$ is our
  preferred model for total functions with Galois sliicing. We
  explore some specific examples in this category in
  \secref{semantic-gps}.
\item The category $\MeetSLat$ has meet-semilattices with top as
  objects and monotone finite meet preserving functions as
  morphisms. The category $\Fam(\MeetSLat)$ provides a model of
  total functions with ``forward derivatives'' only.
\item The category $\JoinSLat$ has join-semilattices with bottom as
  objects and monotone finite join preserving functions as
  morphisms. The category $\Fam(\JoinSLat^\op)$ provides a model of
  total functions with ``backwards derivatives'' only.
\end{enumerate}

We can get an analogous result to \propref{embed-manifolds} for
L-posets and stable maps:

@@

The categories $\Fam(\cat{C})$ are the free coproduct completions of
categories $\cat{C}$, so they have all coproducts:

@@

Using the infinitary coproducts and finite products, we can construct
a wide range of other useful semantic models of datatypes in
$\Fam(\cat{C})$. For example, lists can be constructed as a coproduct
\begin{displaymath}
  \List(X) = \coprod_{n \in \mathbb{N}} X^n
\end{displaymath}
where $X^0 = 1$ (the terminal object) and $X^{n+1} = X \times X^n$.

Our category of interest, $\Fam(\LatGal)$ has coproducts and finite
products, because $\LatGal$ has products. Similarly for
$\Fam(\MeetSLat)$. As we shall see below, the products in $\LatGal$
(and $\MeetSLat$ and $\JoinSLat)$ are also coproducts, which is
essential to obtaining cartesian closure.

@@

For cartesian closure of the categories $\Fam(C)$ that we are
interested in, we rely on the following theorem of \citet{nunes2023},
specialised from their setting with the general Grothendieck
construction to $\Fam(\cat{C})$. This relies on the definition of
\emph{biproducts}, which we discuss below in \secref{biproducts}.

\begin{theorem}[\cite{nunes2023}]
  \label{thm:fam-closed}
  \AGDA.  If $\cat{C}$ has biproducts (\defref{biproducts}) and all
  products, then $\Fam(\cat{C})$ is cartesian closed\footnote{More
    precisely, if $\cat{C}$ has coproducts then we have a monoidal
    product on $\Fam(\cat{C})$ which is closed by this
    construction. When these coproducts are in fact biproducts, we get
    cartesian closure.}. On objects, the internal hom can be given by:
  \begin{displaymath}
    (X, \partial X) \to (Y, \partial Y) = (\Pi_{x : X}. \Sigma_{y : Y}. \cat{C}(\partial X(x), \partial Y(y)), \lambda f. \Pi_{x : X}. \partial Y(\pi_1(f\, x)))
  \end{displaymath}
\end{theorem}

The $\Set$-component of $(X, \partial X) \to (Y, \partial Y)$ consists
of exactly the morphisms of $\Fam(\cat{C})$, rephrased into a single
object. When $\cat{C} = \FinVect$, these are functions with an
associated linear map at every point, and when $\cat{C} = \LatGal$,
these are functions with an associated Galois connection at every
point. A tangent to a function is then defined to be a mapping from
points in the domain to tangents in the codomain along the function.

The category $\MeetSLat$ satisfies the hypoetheses of
\thmref{fam-closed}, so:
\begin{corollary}
  $\Fam(\MeetSLat)$ is cartesian closed and has all coproducts.
\end{corollary}
Unfortunately, neither of the $\LatGal$ or $\FinVect$ satisfy the
hypotheses of this theorem, because neither of them have infinite
products. We will consider ways to rectify this below in
\secref{fixing-completeness}.

\begin{remark}
  \label{rem:hermida-exponentials}
  There is another construction of internal homs on $\Fam(\cat{C})$
  arising from the use of fibrations for categorical logical
  relations, due to \citet[Corollary 4.12]{hermida99}. If we assume
  that $\cat{C}$ is itself cartesian closed and has all products, then
  we could construct an internal hom as:
  \begin{displaymath}
    (X, \partial X) \to (Y, \partial Y) = (X \to Y, \lambda f. \Pi_{x : X}.\,\partial X(x) \to \partial Y(f\,x))
  \end{displaymath}
  However, for the purposes of modelling differentiable programs, this
  is fatally flawed in that neither of $\LatGal$ nor $\FinVect$ are
  cartesian closed, and there is no way of making them so without
  losing the property of being able to conjunct or add tangents, as we
  shall see below. We will implicitly use Hermida's construction in
  our definability proof in \secref{definability}, where we use a
  logical relations argument to show that every morphism definable in
  the higher order language is also first-order definable.
\end{remark}

@@

Loosely stated, biproducts are objects that are both products and
coproducts. The concept can be defined in any category, as shown by
\citet{karvonen20}, but for our purposes it will be more convenient to
use the shorter definition in categories enriched in commutative
monoids:
\begin{definition}
  A category $\cat{C}$ is enriched in $\CMon$, the category of
  commutative monoids, if every homset $\cat{C}(X,Y)$ is a commutative
  monoid with $(+,0)$ and composition is bilinear:
  \begin{displaymath}
    f \comp \zero = 0 = \zero \comp f
  \end{displaymath}
  \begin{displaymath}
    (f + g) \comp h = (f \comp h) + (g \comp h) \qquad
    h \comp (f + g) = (h \comp f) + (h \comp g)
  \end{displaymath}
\end{definition}
In any $\CMon$-category we can define what it means to be the
biproduct of two objects:
\begin{definition}
  \label{def:biproducts}
  In a $\CMon$-category a biproduct is an object $X \biprod Y$
  together with morphisms

@@

\begin{example}
  The following are $\CMon$-enriched and have finite products, and
  hence biproducts:
  \begin{enumerate}
  \item In $\FinVect$, morphisms are linear maps and so can be added
    and have a zero map. Finite products are given by cartesian
    products of the underlying sets, with the vector operations
    defined pointwise.
  \item In $\LatGal$, right adjoints are summed using meets and left
    adjoints are summed using joins. The zero maps are given by the
    constantly $\top$ and constantly $\bot$ functions
    respectively. Products are given by the cartesian product of the
    underlying set and the one-element lattice for the
    terminal/initial/zero object.
  \item $\MeetSLat$ and $\JoinSLat$ are both $\CMon$-enriched and have
    finite products for similar reasons to $\LatGal$.
  \end{enumerate}
\end{example}

\begin{remark}
  Categories with zero objects cannot be cartesian closed without
  being trivial in the sense of having exactly one morphism between
  every pair of objects because
  $\cat{C}(X, Y) \cong \cat{C}(1 \times X,Y) \cong \cat{C}(0 \times
  X,Y) \cong \cat{C}(0,X \to Y) \cong 1$. Consequently, we cannot
  apply the alternative construction of exponentials described in
  \remref{hermida-exponentials}.
\end{remark}

@@

\begin{proposition}
  \label{prop:ho-embedding}
  The functor
  $\HoEmbed : \Fam(\LatGal) \to \Fam(\MeetSLat \times \JoinSLat^\op)$ is defined
  on objects as
  $\HoEmbed(X, \partial X) = (X, \lambda x. (\partial X(x), \partial
  X(x)))$. This functor is faithful and preserves coproducts and
  finite products.
\end{proposition}

With this embedding functor, we see in
\propref{first-order-type-interp} that the interpretation of
first-order types will be the same up to isomorphism in
$\Fam(\LatGal)$ and $\Fam(\MeetSLat \times \JoinSLat^\op)$, as long as
we interpret the base types as objects in $\Fam(\LatGal)$. At
higher-order, however, the meet-semilattice and join-semilattice sides
of the interpretation will diverge, and it is no longer clear that the
interpretation of programs using higher-order functions internally
will result in Galois connections. In \secref{definability} we will
see that every program with first-order type (even if it uses
higher-order functions internally) does in fact have an interpretation
definable in $\Fam(\LatGal)$.

@@

Our thesis is that $\Fam(\LatGal)$ is a suitable setting for
interpreting first-order programs for \GPS. The above discussion has
been somewhat abstract, so we now consider some examples in the
category $\Fam(\LatGal)$ and how they relate to \GPS.

\todo{Fix a notation for the forwards and backwards maps}

Spelt out in full, it has as objects $(X, \partial X)$, all pairs of a
set $X$ and and for every $x \in X$, a bounded lattice
$\partial X(x)$, and as morphisms
$(X, \partial X) \to (Y, \partial Y)$, all pairs $(f, \partial f)$ of
a function $f: X \to Y$ and for every $x \in X$, a Galois connection
$\partial f(x): \partial X(x) \to \partial Y(f(x))$. The reindexing
along $f$ in the composition
$(g, \partial g) \comp (f, \partial f) = (g \comp f, \reindex{\partial
  g}{f} \comp \partial f)$ selects the appropriate lattice of
approximations and Galois connection at each point, using the baseline
(unapproximated) function $f$.

\subsubsection{Unapproximated Functions}

$\LatGal$ has a terminal (also zero) object $\mathbb{1}$, so there is
a functor $\Disc : \Set \to \Fam(\LatGal)$ that maps a set $X$ to
$(X, \lambda x. \mathbb{1})$ and functions $f$ to morphisms
$(f, \lambda\_.\,\mathrm{id}_{\mathbb{1}})$. This functor preserves
products and coproducts. Therefore, we can take any sets and functions
of interest for modelling primitive types and operations of a
programming language and embed, albeit without any interesting
approximation information.

@@

The operation of adding a new bottom element to a bounded lattice
forms part of a monad $L$ on $\LatGal$. This monad extends to a
(strong) monad $\Lift$ on $\Fam(\LatGal)$ with
$\Lift(X, \partial X) = (X, L \circ \partial X)$. The monad $\Lift$
does not affect the points of the original object, but adds a new
minimum approximation. This monad allows us to construct morphisms in
$\Fam(\LatGal)$ that track information about how values are used
similar to the examples of cm and stable functions we saw in
\secref{approx-as-tangents}.

\newcommand{\Bool}{\mathrm{Bool}}

Let $\Bool = \Disc(\{\mathsf{tt},\mathsf{ff}\})$ be the
(unapproximated) embedding of the booleans and
$\mathrm{or} : \Bool \times \Bool \to \Bool$ be the (unapproximated)
boolean or function. The, using a Moggi-style let notation
\cite{notions-of-computation} for morphisms constructed using the
Monad structure of $\Lift$, we can reproduce the functions
$\mathrm{strictOr}$ and $\mathrm{shortCircuitOr}$ functions from
\exref{strict-short-circuit} (we also assume an if-then-else operation
on booleans, definable from the fact that $\Fam(\LatGal)$ has
coproducts and $\Disc$ preserves them). Both of these expressions
define morphisms $\Lift(\Bool) \times \Lift(\Bool) \to \Lift(\Bool)$
in $\Fam(\LatGal)$:
\begin{displaymath}
  \begin{array}{lcl}
    \mathrm{strictOr}(x,y)&=&\mathrm{let}\,b_1 \Leftarrow x\,\mathrm{in}\,\mathrm{let}\,b_2 \Leftarrow y\,\mathrm{in}\,\eta(\mathrm{or}(b_1,b_2)) \\
    \mathrm{shortCircuitOr}(x,y)&=&\mathrm{let}\,b_1 \Leftarrow x\,\mathrm{in}\,\mathrm{if}\,b_1\,\mathrm{then}\,\eta(\mathsf{tt})\,\mathrm{else}\,y
  \end{array}
\end{displaymath}
Examining the morphisms so defined in $\Fam(\LatGal)$, we can see
that, in the $\Set$ component, they are both exactly the normal
boolean-or operation. However, they have different approximation
behaviour, reflecting the different ways that they examine their
inputs. Let us write $\top,\bot$ for the elements of the approximation
lattice at each point of $\Lift(\Bool)$, then applying the reverse
derivative at the point $(\mathsf{tt},\mathsf{tt})$ to the tangent
$\top$ reveals which of the inputs contributed to the output for each
function:
\begin{displaymath}
  \begin{array}{lcl}
    (\partial \mathrm{strictOr})_r(\mathsf{tt},\mathsf{tt})(\top) &=& (\top, \top)  \\
    (\partial \mathrm{shortCircuitOr})_r(\mathsf{tt},\mathsf{tt})(\top) &=& (\top, \bot)
  \end{array}
\end{displaymath}
In comparison to the categories $\CM$ and $\Lposet$ from
\secref{approx-as-tangents}, we have retained the usage information in
the forward and reverse tangents, but we also accurately model
totality of the functions. That is, the constantly $\bot$ function is
also present in both $\CM$ and $\Lposet$, but is not expressible in
$\Fam(\LatGal)$.

An analogue of the $\mathrm{parallelOr}$ function from
\exref{parallel-or} is not definable in $\Fam(\LatGal)$. We would have
to have
$(\partial \mathrm{parallelOr})_f(\mathsf{tt},\mathsf{tt})(\top,\bot)
= (\partial \mathrm{parallelOr})_f(\mathsf{tt},\mathsf{tt})(\bot,\top)
= \top$ to reflect the desired property that either of the inputs
being $\mathsf{tt}$ is enough to determine the output. We also must
have
$(\partial \mathrm{parallelOr})_f(\mathsf{tt},\mathsf{tt})(\bot,\bot)
= \bot$, to reflect the fact that we will get no information in the
output if we required that neither of the inputs is examined. However,
this means that
$(\partial \mathrm{parallelOr})_f(\mathsf{tt},\mathsf{tt})$ will not
presere meets because $(\top,\bot) \sqcap (\bot,\top) = (\bot,\bot)$
but $\top \neq \bot$.

An analogue of the $\mathrm{gustave}$ function from
\exref{parallel-or} is definable in $\Fam(\LatGal)$, but not using the
lifting monad structure as we could for $\mathrm{strictOr}$ and
$\mathrm{shortCircuitOr}$.

\bob{We don't get a theorem like ``if we didn't input $i$ to obtain
  the output, then we could replace it with any other value.'' How
  would we extend $\Fam(\LatGal)$ to do this?}

@@
